Reconstructing Propositional Proofs in Type Theory

### 2017-11-21 paper/sections/conclusions-and-future-work.tex
[ paper ] lots of changes.
[ paper ] lots of fixes and additions.

We presented a reliable proof-reconstruction in type theory
for the propositional fragment of the \Metis prover. We provided for
each \Metis inference rule a formal description in type theory
following a syntactical approach.
This formalizations are mainly exposed in
Section~\ref{ssec:emulating-inferences}.

@@

The reconstruction approach in this study was designed to use
only syntactical aspects of the logic.
This decision was in the beginning a drawback
since it demands more detailed proofs, a
description of every transformation or deduction step performed by
the prover, which is rarely included in the output of these programs,
see Section~\ref{ssec:emulating-inferences}.
Nevertheless, we chose that syntactical treatment instead of using
semantics to extend this work towards the support of first-order logic
or other non-classical logics.
For predicative logic, recall satisfiability is undecidable and
its syntactical aspect plays an important role to reconstruct proofs.

@@

\begin{itemize}
\item extend the proof-reconstruction presented in this paper to
  \begin{itemize}
    \item support the proposition fragment with equality of \Metis.
    \item support other \ATPs for propositional logic like \name{EProver} or \name{Z3}. This development can be carried out by following the \name{EProver}
description on Kanso's Ph.D. thesis~\cite{Kanso2012}.
    \item support \Metis first-order proofs.
  \end{itemize}

\item use a shallow embedding instead the deep embedding
implemented in~\cite{AgdaProp} to reconstruct proofs.
\item improve some functions in Section~\ref{ssec:emulating-inferences}
\begin{itemize}
  \item by investigating the consequences of removing the \clausify
inference rule by the \canonicalize rule.
  \item by increasing the coverage of the \simplify rule. Since is fairly
complex its implementation in the \Metis source code, some cases could be omitted in Section~\ref{sssec:simplify}.
\end{itemize}

>>>>

We presented a proof-reconstruction approach in type theory for the
propositional fragment of the \Metis prover. We provided for each
\Metis inference rule a formal description in type theory following a
syntactical approach. This formalizations are mainly exposed in
Section~\ref{ssec:emulating-inferences}.

@@

The reconstruction approach in this study was designed to use
only syntactical aspects of the logic.
This decision was in the beginning a drawback
since it demands more detailed proofs, a
description of every transformation or deduction step performed by
the prover, which is rarely included in the output of these programs,
see Section~\ref{ssec:emulating-inferences}.
Nevertheless, we chose that syntactical treatment instead of using
semantics to extend this work towards the support of first-order
logic or other non-classical logics.
For first-order logic, recall satisfiability is undecidable and
its syntactical aspect plays an important role to reconstruct proofs.

@@

\begin{itemize}
\item extend the proof-reconstruction presented in this paper to
  \begin{itemize}
    \item support \Metis inference rules with equality (\eg, equality).
    \item support other \ATPs for propositional logic like \name{EProver} or \name{Z3}. This development can be carried out by following the \name{EProver}
description on Kanso's Ph.D. thesis~\cite{Kanso2012}.
    \item support \Metis first-order proofs.
  \end{itemize}

% \item use a shallow embedding instead the deep embedding
% implemented in~\cite{AgdaProp} to reconstruct proofs.
\item improve some functions in Section~\ref{ssec:emulating-inferences}
\begin{itemize}
  \item by investigating the consequences of removing the \clausify
inference rule by the \canonicalize rule.
  \item by increasing the coverage of the \simplify rule. Since is fairly
complex its implementation in the \Metis source code, some cases could be omitted in Section~\ref{sssec:simplify}.
\end{itemize}

### 2017-11-21 paper/sections/proof-reconstruction/clausify.tex
[ paper ] lots of fixes and additions.

The \clausify rule is another alternative way to transform a formula
into its clausal normal form but without any simplification since
remember this conversion is taking into account by the \canonicalize
rule. It is important to notice that this kind of conversion between
one formula to its clausal normal form is not unique, and \Metis has
a customized approach to perform that transformation. Hence, we
perform a reordering of the conjunctive normal form for the source
formula.

>>>>

The \clausify rule is an alternative rule to transform a formula into
its clausal normal form but without performing simplifications with
tautologies or definitions. Recall, this kind of conversion was
addressed by the \canonicalize rule. It is important to notice that
this kind of conversions between one formula to its clausal normal
form are not unique, and \Metis has customized approaches to perform
that transformations. Therefore, we perform a reordering of the
conjunctive normal form given by the $\fcnf$ function defined in
Lemma~\ref{lem:cnf} with the $\freorder_{∧∨}$ function from
Lemma~\ref{lem:reorder-and-or} to the input formula of the rule.

### 2017-11-21 paper/sections/proof-reconstruction/resolve.tex
[ paper ] lots of fixes and additions.

First, we define the \fassoc$_{⬓}$ function in~\eqref{eq:fassoc} to
convert a disjunction or a conjunction into its right-associative
form. The square symbol (⬓) can be the conjunction symbol or the
disjunction symbol. We use $\fassoc_{∧}$ to convert conjunctions and
$\fassoc_{∨}$ for disjunctions.

@@

Now, we are ready to reconstruct the \resolve rule using
Lemma~\ref{lem:reorder-and-or}.
The \resolve rule is a generalization of the resolution theorem.
This rule has three arguments: a positive literal $ℓ$ to apply the
resolution theorem and two derivations.

\begin{mydefinition}
A \emph{literal} is an atom (positive literal) or a negation
of an atom (negative literal).
\end{mydefinition}

@@

\begin{mainlemma}
  \label{lem:rsol}
  If $Γ ⊢ φ$ then $Γ ⊢ \frsol~φ$, where
  \begin{equation}
    \label{eq:rsol}
    \begin{aligned}
    &\frsol : \Prop \to \Prop\\
    &\begin{array}{ll}
      \frsol~φ &=
        \begin{cases}
          φ₂,      &\text{ if }φ ≡ (φ₁ ∨ φ₂) ∨ (¬φ₁ ∨ φ₂);\\
          φ₂ ∨ φ₄, &\text{ if }φ ≡ (φ₁ ∨ φ₂) ∨ (¬φ₁ ∨ φ₄);\\\
          φ, &\text{ otherwise.}
        \end{cases}
      \end{array}
      \end{aligned}
\end{equation}
\end{mainlemma}

>>>>

First, we define the $\fassoc_{⬓}$ function in~\eqref{eq:fassoc} to
convert a disjunction or a conjunction into its right-associative
form. The square symbol (⬓) can be the conjunction symbol or the
disjunction symbol. We use $\fassoc_{∧}$ to convert conjunctions and
$\fassoc_{∨}$ for disjunctions.

@@

Now, we are ready to reconstruct the \resolve rule using
Lemma~\ref{lem:reorder-and-or}. As we see in the following, the
\resolve rule is the \Metis version of the resolution theorem. This
rule takes into account, two propositions that contain a positive
literal $ℓ$ and its negation $¬ ℓ$ respectively. Then, it produces the
\emph{resolvent}, a disjunction of two propositions: the first
proposition after removing the literal $ℓ$ and the second proposition
after removing its negation $¬ ℓ$.

\begin{mydefinition}

A \emph{literal} is an propositional variable (positive literal) or a
negation of an propositional variable (negative literal).

\end{mydefinition}

@@

\begin{mainlemma}
  \label{lem:rsol}
  If $Γ ⊢ φ$ then $Γ ⊢ \frsol~φ$, where
  \begin{equation}
    \label{eq:rsol}
    \begin{aligned}
    &\frsol : \Prop \to \Prop\\
    &\begin{array}{ll}
      \frsol~φ &=
        \begin{cases}
          φ₂,      &\text{ if }φ ≡ (φ₁ ∨ φ₂) ∧ (¬ φ₁ ∨ φ₂);\\
          φ₂ ∨ φ₄, &\text{ if }φ ≡ (φ₁ ∨ φ₂) ∧ (¬ φ₁ ∨ φ₄);\\\
          φ, &\text{ otherwise.}
        \end{cases}
      \end{array}
      \end{aligned}
\end{equation}
\end{mainlemma}
